---
title: "Wealth scores and quintiles with rEquityTool"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Wealth scores and quintiles with rEquityTool}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 84)
```

```{r}
library(rEquityTool)
```

## What this package does

The [EquityTool](https://equitytool.org) is a short asset questionnaire that
places a household in a wealth quintile. It is built by simplifying the full
Demographic and Health Survey (DHS) or Multiple Indicator Cluster Survey (MICS)
wealth index down to the fewest questions that still reproduce it, validated to
a kappa of at least 0.75.

Metrics for Management publish, for each country, an Excel workbook containing
the questionnaire, a score for every response option, and the quintile
cut-points. Applying them is arithmetic: look up a score for each answer, add
them up, compare the total with four thresholds.

Doing that by hand is where it goes wrong. `rEquityTool` reads the workbook,
checks your data, and does the arithmetic.

## A worked example

The package ships a **fabricated** workbook so that everything here runs
without redistributing EquityTool material. Every number in it is made up.

```{r}
et_example()
```

For real work, download your country's workbook from
<https://equitytool.org/countries/> and pass the path to `et_parse_workbook()`.

```{r}
tool <- et_parse_workbook(et_example())
tool
```

The questionnaire, with the response option each code refers to:

```{r}
head(et_questions(tool), 6)
```

and the scores behind it:

```{r}
head(et_scores(tool), 6)
et_cutoffs(tool)
```

## Check your data before you score

The most common error is a mis-coded response option, and it fails *silently*:
an unrecognised code becomes `NA`, the household drops out, and the quintiles
that remain still look perfectly reasonable. Check first.

```{r}
data(demo_survey)
et_validate(demo_survey, tool)
```

`demo_survey` is deliberately imperfect: twelve households never answered `Q3`,
seven have `Q5` coded `9` for "don't know", and four have `Q7` coded `0`. Those
are the two failure modes to look for — genuinely missing answers, and codes
that are not in the tool's list.

If your columns are not named `Q1`, `Q2`, ..., map them. A partial mapping is
fine: name only the ones that differ.

```{r}
survey2 <- demo_survey
names(survey2)[names(survey2) == "Q1"] <- "has_electricity"

head(et_score(survey2, tool, scope = "national",
              vars = c(Q1 = "has_electricity")), 3)
```

## Choosing a scope

`scope` decides **what the quintiles are relative to**. This is a substantive
choice, not a technical one, and it should be stated in your methods section.

### National quintiles

`scope = "national"` uses the published cut-points. Quintile 1 is the poorest
fifth *of the country* at the time of the source survey.

```{r}
nat <- et_score(demo_survey, tool, scope = "national")
table(nat$quintile, useNA = "ifany")
```

The groups are **not** equal in size, and they are not supposed to be. If you
survey a poor district, most of your households land in the bottom quintiles.
That is the finding. A published survey of conflict-affected townships in
southeast Myanmar, for instance, found 59% of households in the poorest
national quintile.

### Urban quintiles

`scope = "urban"` compares urban respondents with the national *urban*
population, which is the right reference for a city survey.

```{r}
table(et_score(demo_survey, tool, scope = "urban")$quintile, useNA = "ifany")
```

### Sample quintiles

Sometimes the question is not "how poor are these households compared with the
country" but "who is worst off *among the people we surveyed*". For that, cut on
your own sample's score distribution.

```{r}
smp <- et_score(demo_survey, tool, scope = "sample", weights = "wt")
table(smp$quintile, useNA = "ifany")
```

Now the groups are approximately equal by construction. Pass `weights` so the
cut-points reflect the population your sample represents rather than the
achieved sample.

The scores are identical across scopes — only the cut-points change:

```{r}
all.equal(nat$score, smp$score)
attr(smp, "cutoffs")
```

**Say which you used.** Sample quintiles are not national quintiles, and
reporting one as the other is a real error: "the poorest quintile in our sample"
and "the poorest national quintile" can be very different groups of people.

You are not restricted to fifths:

```{r}
tertiles <- et_score(demo_survey, tool, scope = "sample", probs = c(1, 2) / 3)
table(tertiles$quintile)
```

## Missing data

The official rule is that a household with **any** missing or out-of-range
answer gets no score at all. `rEquityTool` follows it:

```{r}
sum(is.na(nat$score))
```

`na_rule = "partial"` will instead sum whatever is available. This is **not**
the official method — a household answering six of eight questions gets a score
that is not comparable with one answering all eight. Use it for diagnostics
only.

```{r}
sum(is.na(et_score(demo_survey, tool, na_rule = "partial")$score))
```

## Rural/urban tools

Fourteen of the published country tools do not have a single national score
set. Instead they score urban and rural respondents on *separate* scales and map
both onto a common national scale with a published linear equation. For those
tools you must say who lives where.

```{r}
ru <- et_parse_workbook(et_example("rural_urban"))
ru
```

```{r}
res <- et_score(demo_survey, ru, scope = "national", area = "residence")
head(res, 3)
```

`area` accepts the EquityTool's own coding (`1` urban, `2` rural), the words
`"urban"`/`"rural"`, or a bare vector. `area_score` holds the pre-rescaling
score. Forgetting `area` is an error rather than a silently wrong answer:

```{r, error = TRUE}
et_score(demo_survey, ru, scope = "national")
```

## Putting it to work

Attach the results to your data with `append = TRUE` and go:

```{r}
scored <- et_score(demo_survey, tool, scope = "national", append = TRUE)

round(100 * prop.table(table(scored$region, scored$quintile), 1), 1)
```

```{r}
# coverage of an outcome across wealth groups -- the input to an equiplot
# or a concentration index
aggregate(improved_water ~ quintile, data = scored, FUN = mean)
```

## Two defects in the published workbooks

The package reads each workbook's scoring rules from the SPSS syntax tab and
then cross-checks them against the score table printed on the *Other software*
tab. Across the 69 country workbooks published as of September 2026, two
disagree.

**Cambodia (DHS 2021).** The national block of the *Other software* tab lists
the `Q8` response options in a different order from the syntax tabs and from its
own urban block. The two documented routes give different scores for the same
household. `rEquityTool` uses the syntax and records the conflict:

```{r, eval = FALSE}
tool <- et_parse_workbook("Cambodia-DHS-2021-public-file-2023-11-14.xlsx")
#> Warning: 2 score(s) differ between the 'SPSS Syntax' tab and the
#> 'Other software' tab. The syntax tab has been used.
et_checks(tool)$mismatches
```

**Djibouti (EDAM 2017).** The urban quintile block opens with
`DO IF NationalScore >=1.090269434498.` followed by `COMPUTE UrbanQuintile =5.`
— the condition names the wrong variable. Running that syntax verbatim assigns
the top urban quintile on the basis of the national score. The workbook's own
table confirms 1.0903 is the urban boundary. Because `rEquityTool` reads the
`COMPUTE` assignment rather than the condition, it gets this right, and flags
it.

Always look at the cross-check before trusting a run:

```{r}
et_checks(tool)$mismatches
et_checks(tool)$cutoff_mismatches
```

## Caveats worth carrying into your write-up

- The EquityTool is a **simplified** index. Agreement with the full DHS/MICS
  wealth index is around kappa 0.75–0.85, so individual households can be
  misclassified by one quintile. It is designed for describing groups, not for
  adjudicating a single household's status.
- Cut-points are anchored to the **national distribution at the time of the
  source survey**. The Myanmar tool, for example, is built on DHS 2015–16. The
  further your fieldwork is from that date, the more the "national quintile"
  label is a comparison with a historical benchmark.
- Urban and rural indices measure position **within** those populations. An
  urban quintile 1 household is not necessarily poorer than a rural quintile 5
  household.
- These are **relative** measures of wealth, not absolute measures of poverty.

## Where the wealth index goes next

The score and quintile from this package are the ranking variable for the usual
equity analyses:

- concentration indices and decomposition — see the
  [`rineq`](https://cran.r-project.org/package=rineq) package;
- equiplots and the slope index of inequality — see
  [`ICEHmeasures`](https://cran.r-project.org/package=ICEHmeasures).

Use `score` rather than `quintile` when you need a continuous ranking variable,
and remember to carry your survey design through into those analyses.
