---
title: "Choosing a wealth quintile scope"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing a wealth quintile scope}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 84)
```

```{r}
library(rEquityTool)
tool <- et_parse_workbook(et_example())
data(demo_survey)
```

`et_score()` has a `scope` argument with four settings. It is the most
consequential choice in the package, and it is a **substantive** one, not a
technical detail: it decides *what your quintiles are relative to*.

| `scope` | Cut-points come from | Quintile 1 is the poorest fifth of... |
|---|---|---|
| `"national"` | The published national index | the country, at the time of the source survey |
| `"urban"` | The published urban index | the national urban population |
| `"rural"` | The published rural index, where one exists | the national rural population |
| `"sample"` | **Your own data** | your respondents |

## The score does not change. The cut-points do.

This is the thing to internalise. All four scopes on the same households:

```{r}
nat <- et_score(demo_survey, tool, scope = "national")
smp <- et_score(demo_survey, tool, scope = "sample", weights = "wt")

all.equal(nat$score, smp$score)
```

The wealth score is a property of the household's assets. The *quintile* is a
statement about where that score sits relative to some reference population.
Changing `scope` changes the reference population, nothing else.

```{r}
rbind(
  national = table(nat$quintile),
  sample   = table(smp$quintile)
)
```

## `"national"` — position in the country

Use this when you want to say where your respondents sit in the national
distribution. This is the default and the right choice for most programme
reporting.

```{r}
round(100 * prop.table(table(nat$quintile)), 1)
```

**The groups will not be equal, and they are not supposed to be.** If you
survey a poor district, most households land in the bottom quintiles. That is
the finding. A published midline survey of conflict-affected townships in
Kayin State, Myanmar found 59.0% of households in the poorest national quintile
and 2.6% in the richest — an informative result that sample-based quintiles
would have concealed entirely.

A common mistake is to see an uneven distribution, assume something is broken,
and switch to `"sample"` to "fix" it. That discards the finding.

## `"urban"` — position among urban residents

Against the national distribution, urban residents mostly look rich, because
they mostly are. If your survey is of a city, that comparison is uninformative:

```{r}
urb_only <- demo_survey[demo_survey$residence == "urban", ]
table(et_score(urb_only, tool, scope = "national")$quintile)
```

The urban index gives a reference population that discriminates within that
group:

```{r}
table(et_score(urb_only, tool, scope = "urban")$quintile)
```

Note that this is not merely a different set of cut-points — the urban index
**weights the assets differently**, because what distinguishes rich from poor
in a city differs from what distinguishes them nationally:

```{r}
s <- et_scores(tool)
merge(
  s[s$scope == "national" & s$option == 1, c("var", "score")],
  s[s$scope == "urban"    & s$option == 1, c("var", "score")],
  by = "var", suffixes = c("_national", "_urban")
)
```

Apply the urban index only to urban respondents. An urban quintile 1 household
is **not** necessarily poorer than a rural quintile 5 household; the two are
measured against different populations.

## `"rural"` — where the tool publishes it

Some tools publish a rural index. Many do not: in the 14 "rural + urban"
country tools the rural score is an *intermediate* quantity that gets rescaled
onto the national scale, and no rural cut-points are published. Ask the tool:

```{r}
et_scopes(tool)
```

Requesting a scope the tool does not support is an error, not a silent
fallback:

```{r, error = TRUE}
et_score(demo_survey, tool, scope = "rural")
```

## `"sample"` — position among your own respondents

Sometimes the question is not "how poor are these households compared with the
country" but "who is worst off **among the people we surveyed**". That is a
legitimate question — it is the right one when you are analysing a gradient
*within* a study population, or when national quintiles leave you with three
households in the top group and no power to say anything about them.

```{r}
table(smp$quintile)
```

The groups are now approximately equal by construction, and the cut-points came
from your data:

```{r}
attr(smp, "cutoffs")
```

### Always pass `weights`

Without weights the cut-points describe your *achieved sample*. With weights
they describe the population your sample was designed to represent. These are
different, sometimes materially:

```{r}
unweighted <- et_score(demo_survey, tool, scope = "sample")
weighted   <- et_score(demo_survey, tool, scope = "sample", weights = "wt")

rbind(unweighted = attr(unweighted, "cutoffs"),
      weighted   = attr(weighted,   "cutoffs"))
```

```{r}
table(unweighted = unweighted$quintile, weighted = weighted$quintile)
```

### You are not limited to fifths

`probs` takes any set of cut probabilities, so tertiles, quartiles and deciles
all work. Useful when a sample is too small to support five groups:

```{r}
tertiles <- et_score(demo_survey, tool, scope = "sample", probs = c(1, 2) / 3)
table(tertiles$quintile)
```

```{r}
deciles <- et_score(demo_survey, tool, scope = "sample", probs = seq(0.1, 0.9, 0.1))
table(deciles$quintile)
```

## How much does the choice actually matter?

A lot. Here is where households end up under the two classifications:

```{r}
table(national = nat$quintile, sample = smp$quintile)
```

Off-diagonal cells are households that would be described completely
differently depending on a single argument.

## Reporting rules

1. **State which scope you used, in the methods.** "Wealth quintiles" is not
   sufficient. Write "national wealth quintiles from the Myanmar EquityTool"
   or "quintiles of the weighted sample wealth score distribution".
2. **Never call sample quintiles national quintiles.** They are different
   groups of people. This is a substantive error, not a presentational one.
3. **If you report national quintiles, say when the reference survey was.**
   Cut-points are anchored to the national distribution at that date.
4. **Do not switch to `"sample"` because the national distribution looks
   lopsided.** That lopsidedness is usually your headline result.

A reasonable default for programme evaluation: report **national** quintiles to
place your population in its country context, and use **sample** quintiles when
you need groups of workable size for within-study gradient analysis. Report
both if they tell different stories — that difference is itself informative.

## See also

- `vignette("myanmar-worked-example")` — the whole workflow with a real
  country workbook.
- `vignette("workbooks")` — what to do when a tool has rural/urban scales or
  internal contradictions.
