---
title: "Working with different country workbooks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with different country workbooks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 84)
```

```{r}
library(rEquityTool)
```

The 69 country workbooks published at <https://equitytool.org/countries/> are
not all the same shape. This article covers the structural variants, how to
find your way around a folder of downloads, and how to check a workbook before
you trust it.

## Finding what you have

If you have downloaded several countries, `et_find_workbooks()` inventories a
folder. With `inspect = TRUE` it opens each file and reports what is in it:

```{r}
dir <- dirname(et_example())
et_find_workbooks(dir, inspect = TRUE)[, c("country", "variant", "n_questions",
                                           "scopes", "issues")]
```

`issues` is the column to read. An empty string means the workbook is
internally consistent; anything else needs attention before you use it.

## The three structural variants

`et_parse_workbook()` reports which variant a tool uses, and `et_score()`
adapts.

| Variant | Countries | What it means |
|---|---|---|
| `national_urban` | 53 | Separate national and urban indices. The common case |
| `rural_urban` | 14 | Separate **rural** and **urban** indices, rescaled onto a national scale |
| `national_only` | 2 | National index only (Argentina, Uruguay) |

### `national_urban` — the common case

```{r}
tool <- et_parse_workbook(et_example())
tool
```

Nothing special is required. Pick a scope and score:

```{r}
data(demo_survey)
table(et_score(demo_survey, tool, scope = "national")$quintile)
```

### `rural_urban` — needs an area indicator

Fourteen tools (Afghanistan, Armenia, Burkina Faso, Chad, Djibouti, Egypt,
Ethiopia, Laos, Niger, Tajikistan, Thailand, Tunisia, Vietnam, Zimbabwe) score
urban and rural respondents on **different scales**, then map both onto a
common national scale with a published linear equation:

```
RuralScore    = sum of the rural item scores
UrbanScore    = sum of the urban item scores   (negated in Burkina Faso)
NationalScore = a_urban + b_urban x UrbanScore    if the household is urban
NationalScore = a_rural + b_rural x RuralScore    if the household is rural
```

The package reads those coefficients out of the workbook and reports them:

```{r}
ru <- et_parse_workbook(et_example("rural_urban"))
ru
```

Because the two scales are different, scoring on the national scale is
impossible without knowing who lives where. Omitting `area` is an error rather
than a wrong answer:

```{r, error = TRUE}
et_score(demo_survey, ru, scope = "national")
```

```{r}
res <- et_score(demo_survey, ru, scope = "national", area = "residence")
head(res, 4)
```

`area_score` is the pre-rescaling value on the household's own scale; `score`
is after rescaling onto the national scale.

`area` is matched leniently — the EquityTool's own coding (`1` urban, `2`
rural), the words, or an abbreviation, in any case:

```{r}
a <- et_score(demo_survey, ru, "national", area = "residence")
b <- et_score(demo_survey, ru, "national",
              area = ifelse(demo_survey$residence == "urban", 1L, 2L))
c3 <- et_score(demo_survey, ru, "national",
               area = toupper(as.character(demo_survey$residence)))
c(identical(a$score, b$score), identical(a$score, c3$score))
```

**These tools publish national and urban cut-points but no rural ones.** The
rural score is an intermediate. Check before assuming:

```{r}
et_scopes(ru)
```

Two of them have further quirks the package handles automatically: **Burkina
Faso** negates the urban score before rescaling, and **Tunisia** has negative
rescaling slopes.

### `national_only`

Argentina and Uruguay publish no urban tool. `et_scopes()` will say so, and
requesting `"urban"` errors.

## Checking a workbook before you trust it

Every EquityTool workbook states its scoring rules **twice**: once as runnable
SPSS and Stata syntax, and once as a printed table on the *Other software* tab.

`rEquityTool` reads the **syntax** — it maps a response *code* directly to a
score, with no ambiguity about option ordering — and cross-checks it against
the printed table. Disagreements are recorded:

```{r}
str(et_checks(tool), max.level = 1)
```

```{r}
et_checks(tool)$mismatches
et_checks(tool)$cutoff_mismatches
et_checks(tool)$mislabelled_conditions
```

Three empty data frames means the workbook agrees with itself. Make this a
habit: it takes one line and it is the only way to catch the problems below.

## Three real defects in the published corpus

Checking all 69 workbooks turned up three problems in the official files. None
is documented by the publisher, and none had been reported at the time of
writing.

### Cambodia DHS 2021 — the two routes disagree

The *national* block of the *Other software* tab lists the `Q8` response
options (floor material) in a **different order** from the syntax tabs, and
from its own urban block:

| Route | Option 1 | Option 2 |
|---|---|---|
| *Other software* tab, national | ceramic tiles → 0.177085 | wood planks → 0.686445 |
| SPSS / Stata / R tabs | 0.686445 | 0.177085 |

A household scored by hand from the printed table gets a different wealth score
than one scored with the supplied syntax. Parsing warns, and records it:

```{r, eval = FALSE}
cam <- et_parse_workbook("Cambodia-DHS-2021-public-file-2023-11-14.xlsx")
#> Warning: 2 score(s) differ between the 'SPSS Syntax' tab and the
#> 'Other software' tab. The syntax tab has been used. See et_checks() for details.

et_checks(cam)$mismatches
#>   var    scope option    syntax other_tab
#> 1  Q8 national      1 0.6864448 0.1770851
#> 2  Q8 national      2 0.1770851 0.6864448
```

### Djibouti EDAM 2017 — the urban cut-point tests the wrong variable

The urban quintile block opens:

```
DO IF NationalScore >=1.090269434498.
COMPUTE UrbanQuintile =5.
```

The condition names `NationalScore` where it should name `UrbanScore`. The
workbook's own table confirms 1.0903 is the **urban** quintile-5 boundary.
Anyone running the published SPSS syntax verbatim assigns the top urban
quintile on the wrong variable.

This is why the package reads cut-points **structurally** — pairing each
threshold with the `COMPUTE <scope>Quintile = k` statement it guards, so the
scope and quintile number come from the assignment rather than from the
condition:

```{r, eval = FALSE}
dji <- et_parse_workbook("Djibouti-EDAM-2017-public-file-2022-10-12.xlsx")

et_checks(dji)$mislabelled_conditions
#>   scope quintile condition_var
#> 1 urban        5      national

subset(et_cutoffs(dji), scope == "urban")
#>   scope quintile      lower
#> 5 urban        5  1.0902694   <- correctly attributed to urban
#> 6 urban        4  0.1376637
#> 7 urban        3 -0.3511468
#> 8 urban        2 -0.9498314
```

### Djibouti EDAM 2017 — a stale country name

The SPSS tab identifies itself as `VietnamMICS2021`, while the Stata tab and
the rest of the file correctly say `DjiboutiEDAM2017` — a copy-paste left in
place. Cosmetic, but it would mislabel your output. The package resolves the
survey name by majority vote across tabs.

## A defensive reading routine

```{r, eval = FALSE}
tool <- et_parse_workbook("your-country-file.xlsx")

# 1. Is it the tool you expected, and what variant?
tool

# 2. Does the workbook contradict itself?
ck <- et_checks(tool)
stopifnot(nrow(ck$mismatches) == 0, nrow(ck$cutoff_mismatches) == 0)

# 3. Which scopes can you actually use?
et_scopes(tool)

# 4. Does your data match the questionnaire?
et_validate(my_survey, tool)

# 5. Only now, score.
et_score(my_survey, tool, scope = "national")
```

## Reporting a problem

If you hit a workbook that contradicts itself, the output of `et_checks()` is
exactly what the publisher needs. Contact Metrics for Management at
<equitytool@m4mgmt.org>, and please also open an issue at
<https://github.com/rEquityTools/rEquityTool/issues> so the finding is recorded
for other users.

## See also

- `vignette("scopes")` — choosing between national, urban, rural and sample
  cut-points.
- `vignette("myanmar-worked-example")` — the whole workflow with a real
  country workbook.
